UKB Simulation Analyses

### 2023-12-06 paper_figs.Rmd
adding gams for extremeness bias tests

source("./R_scripts/gamlss_helper_funs.R")
```

@@

### LOAD ###
pheno_list <- readRDS(file="R_scripts/pheno_list.rds")
vol_list_global <- readRDS(file="R_scripts/vol_list_global.rds")
vol_list_regions <- readRDS(file="R_scripts/Vol_list_regions.rds")
sa_list <- readRDS(file="R_scripts/SA_list.rds")
ct_list <- readRDS(file="R_scripts/CT_list.rds")

@@

#ukb data from ohbm_results.Rmd
cent_err.p <- readRDS("ukb_perm_csvs/centile_errors.RDS")
z_diffs2 <- readRDS("ukb_basic/z_diffs.RDS")
cent.diffs3 <- readRDS("ukb_basic/cent_diffs.RDS")
ukb_data <- fread("data/ukb_to_model/ukb_CN_data_simsites.csv")
#cent_abs.diffs <- readRDS("data/centile_abs_diffs.rds")
```

>>>>

source("./cubic_scripts/R_scripts/gamlss_helper_funs.R")
```

@@

### LOAD ###
pheno_list <- readRDS(file="cubic_scripts/R_scripts/pheno_list.rds")
vol_list_global <- readRDS(file="cubic_scripts/R_scripts/vol_list_global.rds")
vol_list_regions <- readRDS(file="cubic_scripts/R_scripts/Vol_list_regions.rds")
sa_list <- readRDS(file="cubic_scripts/R_scripts/SA_list.rds")
ct_list <- readRDS(file="cubic_scripts/R_scripts/CT_list.rds")

@@

#ukb data from ohbm_results.Rmd
cent_err.p <- readRDS("data/ukb_perm_csvs/centile_errors.RDS")
z_diffs2 <- readRDS("ukb_basic/z_diffs.RDS")
cent.diffs3 <- readRDS("ukb_basic/cent_diffs.RDS")
ukb_data <- fread("data/ukb_to_model/ukb_CN_data_simsites.csv")
#cent_abs.diffs <- readRDS("data/centile_abs_diffs.rds")
```

### 2023-12-06 ukb_sim.site_analyses.Rmd
adding gams for extremeness bias tests

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(gamlss, gamlss.cens, gamlss.dist, gamlss.mx, gamlss.tr, ggplot2, tidyverse, ggpubr, skimr, gamlss.data, data.table, nlme, devtools, slider, tidymodels, broom.mixed, ggseg, cowplot, lsr, paletteer, ggsegExtra, ggseg3d)

@@

summary(sigma.df$beta_ratio). #method 1 male effect on var
summary(combined_var$sex_std) #method 2 male effect on var
```

@@

@@

No shift in abs. differences, looking at overall diff.
```{r}
#get centile errors for each subject - not abs val
cent.diffs3 <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "_refA")) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ . - first(.), .names = "diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop ref
saveRDS(cent.diffs3, file="ukb_basic/cent_diffs.RDS")
```

```{r, eval=FALSE}
cent.diffs3 %>%
  #dplyr::filter(dataset == "cf.gamlss") %>%
  pivot_longer(starts_with("diff_"), names_to="pheno", values_to="diff", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_))) %>%
  ggplot() +
  geom_density(aes(x=diff, fill=dataset), alpha=0.4) +
  facet_wrap(~pheno_cat) +
  theme(legend.position="none")
```

@@

Other methods are fairly consistently underestimating male centiles and overestimating female centiles.

@@

T test of mean diff. between sex w/in each dataset
```{r}
t.test.by.sex <- mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

@@

@@

@@

#### OHBM Abstract

rerunning stats above w/o ref-site combat versions (to account for different FDR correction) per Andrew's suggestion

Test

Test for differences in centile offset distributions
```{r}
mean.diffs.long_filt <- mean.diffs.long %>%
  filter(!str_detect(dataset, "_refA")) %>%
  filter(dataset != "ukb_CN")

summary(aov(centile_abs.diff ~ dataset, data=mean.diffs.long_filt))

attach(mean.diffs.long_filt)
pairwise.t.test(x=centile_abs.diff, g=dataset, p.adj = "bonferroni")
detach()
```

Testing for differences in Z offset distributions

```{r}
mean.z.long_filt <- mean.z.long %>%
  filter(!str_detect(dataset, "_refA"))

summary(aov(z_abs.diff ~ dataset, data=mean.z.long_filt))

attach(mean.z.long_filt)
pairwise.t.test(x=z_abs.diff, g=dataset, p.adj = "fdr")
detach()
```

Test for sex-biases in centile offsets

Test for sex-biases

```{r}
mean.z.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  filter(!str_detect(dataset, "_refA")) %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(z_diff ~ sex)$statistic,
            p_value = t.test(z_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  filter(!str_detect(dataset, "_refA")) %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))
```

##### Plots

Beta weight? or proportion?

```{r}
beta.plot2 <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss"))) %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = var_est)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Cork") +
  ggtitle("Beta weight for effect of male sex on variance")

beta.plot2

ggsave("male_variance_beta2.jpeg", beta.plot2, width=10, height=10)
```

```{r}
sigma.aseg_all <- sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  dplyr::select(!label) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss")),
         tissue_class = as.factor(case_when(pheno == "GMV" ~ "cGM",
                                            pheno == "Ventricles" ~ "CSF",
                                            pheno == "WMV" ~ "White Matter",
                                            pheno == "sGMV" ~ "sGM",
                                            TRUE ~ NA)),
         var_est = case_when(sig.sum_bf.corr == TRUE ~ var_est,
                             TRUE ~ NA)) #just keep significant regions

sigma.aseg_all <- left_join(sigma.aseg_all, aseg_cerebrum$data, by = c("tissue_class"))

sigma.aseg_all %>%
  group_by(dataset) %>%
  ggplot() +
  geom_brain(atlas = aseg_cerebrum, 
             #position = position_brain(hemi ~ side),
             aes(fill = var_est),
             alpha=0.8) +
  #scale_fill_gradient2() +
  facet_wrap(~dataset) + 
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Beta weight for effect of male sex on variance")
```

```{r}
proportion_beta_plot2 <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss"))) %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = beta_ratio)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=15) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Proportion of effect of male sex on variance")

proportion_beta_plot2
ggsave("male_variance_prop.beta2.jpeg", proportion_beta_plot2, width=10, height=10)
```

Centile plot

```{r}
#Define scales
scales_y <- list(
  `Global Volume` = scale_y_continuous(limits = c(0, 4)),
  `Regional CT` = scale_y_continuous(limits = c(0, 70)),
  `Regional SA` = scale_y_continuous(limits = c(0, 70)),
  `Regional Volume` = scale_y_continuous(limits = c(0, 70))
)

cent.diffs.plot2 <- mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  filter(!str_detect(dataset, "_refA")) %>%
  mutate(centile_abs.diff = centile_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_histogram(aes(x=centile_abs.diff, fill=dataset), color="black", alpha=0.4, position="identity") +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma()) + 
  facet_wrap(~pheno_cat) +
  #facetscales::facet_grid_sc(pheno_cat~.,
                #scales = list(y = scales_y)) +
  #paletteer::scale_fill_paletteer_d("Polychrome::light") +
  ggtitle("Mean absolute differences in centile scores across \nComBat processing pipelines for each phenotype")

cent.diffs.plot2
#ggsave("mean_centile_diffs2.jpeg", cent.diffs.plot2, width=10, height=10)
```

>>>>

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(gamlss, gamlss.cens, gamlss.dist, gamlss.mx, gamlss.tr, ggplot2, tidyverse, ggpubr, skimr, gamlss.data, data.table, nlme, devtools, slider, tidymodels, broom.mixed, ggseg, cowplot, lsr, paletteer, ggsegExtra, ggseg3d, mgcv)

@@

summary(sigma.df$beta_ratio) #method 1 male effect on var
summary(combined_var$sex_std) #method 2 male effect on var
```

@@

**check w andrew**

```{r, warning=FALSE, message=FALSE}
#within each phenotype, iterate paired wilcox rank test between each combo of combat dfs

@@

No shift in abs. differences, looking at overall diff.

```{r}
#get centile errors for each subject - not abs val
cent.diffs3 <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "_refA")) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ . - first(.), .names = "diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop ref
saveRDS(cent.diffs3, file="ukb_basic/cent_diffs.RDS")
```

Look at w/in subject maximum error across all feature's centiles & compare across harmonization methods

@@

all methods are fairly consistently underestimating male centiles and overestimating female centiles.

@@

test of mean diff. between sex w/in each dataset - again need to update w nonparametric method **ask andrew about this**
```{r}
t.test.by.sex <- mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

@@

t tests of sex-biases in centiles w/in each feature **ask andrew about this**

@@

### Extremeness Bias

Does a person's centile error depend on their true centile value (e.g., very low or very high centiles more or less likely to have errors)

do some plotting

```{r}
ggplot(cent.diffs3, aes(x=GMV, y=diff_GMV, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  facet_wrap(~dataset)

ggplot(cent.diffs3, aes(x=WMV, y=diff_WMV, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  facet_wrap(~dataset, scales="free")
```

Testing on MAGNITUDE of error, not the errors themselves!

```{r}
ggplot(cent.diffs2, aes(x=GMV, y=diff_GMV, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  facet_wrap(~dataset)

ggplot(cent.diffs2, aes(x=WMV, y=diff_WMV, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  facet_wrap(~dataset, scales="free")
```

*ask andrew about how best to test - would GAM be too flexible? use polynomials?*

@@

### Extremeness Bias

Plot error

```{r}
ggplot(z_diffs2, aes(x=GMV.z, y=diff_GMV.z, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth(method="lm") +
  facet_wrap(~dataset)

ggplot(z_diffs2, aes(x=WMV.z, y=diff_WMV.z, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth(method="lm") +
  facet_wrap(~dataset)
```

plot magnitude error (abs val)

```{r}
ggplot(z_diffs, aes(x=GMV.z, y=abs.diff_GMV.z, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  facet_wrap(~dataset)

ggplot(z_diffs, aes(x=WMV.z, y=abs.diff_WMV.z, color=sex)) +
  geom_point(alpha=0.1) +
  geom_smooth(method="lm") +
  facet_wrap(~dataset)
```

Try using gam to test associations (start with just GMV)

```{r}
m1 <- gam(abs.diff_GMV.z ~ s(GMV.z, by=dataset) + dataset, data=z_diffs)
summary(m1)

#test for interactive sex effects - split df by dataset and smooth by sex
for (dset in c("cf", "cf.gam", "cf.gamlss", "cf.lm")) {
  print(dset)
  df <- z_diffs %>%
    dplyr::filter(dataset == dset)

  m2 <- gam(abs.diff_GMV.z ~ s(GMV.z, by=as.factor(sex)) + as.factor(sex) + s(GMV.z), data=df)
  print(summary(m2))
}
```

Pivot data long and run gam to test assoc. across all features
```{r}
dataset_test_list <- c("cf", "cf.gam", "cf.gamlss", "cf.lm")
pheno_list.z <- paste0(pheno_list, ".z")

z_diffs.long <- z_diffs %>%
  dplyr::filter(dataset %in% dataset_test_list) %>%
  mutate(dataset = droplevels(dataset)) %>%
  dplyr::select(c(participant, sex, age_days, dataset, Source_File), starts_with("abs.diff_")) %>%
  pivot_longer(starts_with("abs.diff_"), names_to="pheno", values_to="z_abs.diff", names_prefix = "abs.diff_")

z_long <- z_diffs %>%
  dplyr::filter(dataset %in% dataset_test_list) %>%
  mutate(dataset = droplevels(dataset)) %>%
  dplyr::select(c(participant, sex, age_days, dataset, Source_File, pheno_list.z)) %>%
  pivot_longer(cols=all_of(pheno_list.z), names_to="pheno", values_to="z_score")

z_long.full <- inner_join(z_diffs.long, z_long)
```

```{r}
gam.ext_m1 <- gam(z_abs.diff ~ s(z_score, by=dataset) + dataset, data=z_long.full)
summary(gam.ext_m1)
```

```{r}
#test for interactive sex effects - split df by dataset and smooth by sex 
for (dset in c("cf", "cf.gam", "cf.gamlss", "cf.lm")) {
  print(dset)
  df <- z_long.full %>%
    dplyr::filter(dataset == dset)

  m2 <- gam(z_abs.diff ~ s(z_score, by=as.factor(sex)) + as.factor(sex) + s(z_score), data=df)
  print(summary(m2))
}
```
